---
title: "anglemania Tutorial"
author:
- name: Aaron Kollotzek
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
  email: aaron.kollotzek@mdc-berlin.de
- name: Vedran Franke
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
- name: Artem Baranovskii
  affiliation:
  - Helmholtz Munich
- name: Altuna Akalin
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{anglemania tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
anglemania is a feature selection package that extracts genes from multi-batch scRNA-seq 
experiments for downstream dataset integration.
The goal is to select genes that carry high biological information and low 
technical noise between the batches. Those genes are extracted from gene pairs 
that have an invariant and extremely narrow or wide angle between their expression vectors.
Conventionally, highly-variable genes (HVGs) or sometimes all genes are used
for integration tasks (https://www.nature.com/articles/s41592-021-01336-8).
While HVGs are a great and easy way to reduce the noise and
dimensionality of the data, we hypothesize that there are better ways to select genes
specifically for integration tasks. HVGs are sensitive to batch effects because the
variance is a function of both the technical and biological variance.
anglemania improves conventional usage of HVGs for integration tasks, 
especially when the transcriptional difference between cell types or cell states 
is subtle (showcased here using simulated data generated using `r Biocpkg("splatter")`
with `de.facLoc` and `de.facScale` set to 0.1, which results in mild differences between 
"Groups").
The package can be used on top of `r Biocpkg("SingleCellExperiment")` or
`r CRANpkg("Seurat")` objects.

Under the hood, anglemania works with file-backed big matrices (FBMs) from
the `r CRANpkg("bigstatsr")` package for fast and memory 
efficient computation.

```{r, load libraries}
suppressPackageStartupMessages({
    library(anglemania)
    library(dplyr)
    library(Seurat)
    options(Seurat.object.assay.version = "v5")
    library(splatter)
    library(SingleCellExperiment)
    library(scater)
    library(scran)
    library(bluster)
    library(batchelor)
    library(UpSetR)
})
```

# Simulation
We simulate a scRNA-seq dataset using Splatter with 4 batches and 3 cell types with subtle differences 
between cell types and rather big batch effects. 
```{r, create simulated data}
batch.facLoc <- 0.3
de.facLoc <- 0.1
nBatches <- 4
nGroups <- 3
nGenes <- 5000
groupCells <- 300

sce_raw <- splatSimulate(
    batchCells = rep(groupCells * nGroups, nBatches),
    batch.facLoc = batch.facLoc,
    group.prob = rep(1 / nGroups, nGroups),
    nGenes = nGenes,
    batch.facScale = 0.1,
    method = "groups",
    verbose = FALSE,
    out.prob = 0.001,
    de.prob = 0.1, # mild
    de.facLoc = de.facLoc,
    de.facScale = 0.1,
    bcv.common = 0.1,
    seed = 42
)
sce <- sce_raw
assays(sce)
```

# Unintegrated data
Here we perform a standard workflow on the unintegrated data. 
When we perform clustering on the unintegrated data and visualize it in a UMAP, 
we can see that the clusters are driven by batch effects rather than cell types.
```{r, fig.cap = "UMAPs of unintegrated data, colored by Batch and Group. The clusters are driven by batch effects.", fig.wide = TRUE, fig.width = 10}
sce_unintegrated <- sce
# Normalization.
sce_unintegrated <- logNormCounts(sce_unintegrated)

# Feature selection.
dec <- modelGeneVar(sce_unintegrated)
hvg <- getTopHVGs(dec, prop = 0.1)

# PCA.
set.seed(1234)
sce_unintegrated <- scater::runPCA(
    sce_unintegrated,
    ncomponents = 50,
    subset_row = hvg
)

# Clustering.
colLabels(sce_unintegrated) <- clusterCells(sce_unintegrated,
    use.dimred = "PCA",
    BLUSPARAM = NNGraphParam(cluster.fun = "louvain")
)

# Visualization.
sce_unintegrated <- scater::runUMAP(sce_unintegrated, dimred = "PCA")
patchwork::wrap_plots(
    plotUMAP(sce_unintegrated, colour_by = "Batch") +
        ggtitle("Unintegrated data, colored by Batch"),
    plotUMAP(sce_unintegrated, colour_by = "Group") +
        ggtitle("Unintegrated data, colored by Group")
)
```



# anglemania
anglemania works on a `r Biocpkg("SingleCellExperiment")` object. 
The function has a few important arguments: 
- `batch_key`: the column in the metadata of the SingleCellExperiment 
object that indicates which batch the cells belong to. This is required to 
distinguish between batches, because we compute the angle between gene pairs 
for each batch.
- `method`: either cosine or spearman
    - this is the method by which the relationship of the gene pairs is measured. 
Default is cosine, which is the cosine similarity between the expression 
vectors of the gene pairs.
- `max_n_genes`: you can specify a maximum number of extracted genes. They are 
sorted by a weighted sum of the ranked mean and standard deviation of the zscores
of the gene pair correlations.

```{r, run anglemania, message = FALSE}
head(colData(sce))
batch_key <- "Batch"
sce <- anglemania(
    sce,
    batch_key = batch_key,
    max_n_genes = 2000
)
```

## extract the anglemania genes from the SCE object
```{r}
anglemania_genes <- get_anglemania_genes(sce)
head(anglemania_genes)
length(anglemania_genes)
```

## select_genes 
Once anglemania was run on the SCE, you can adjust the number of genes you want
to select using the `select_genes()` function. It has an argument score_weights,
which let's you decide how much weight to set to the mean and standard
deviation of the zscores. The default which is used in the `anglemania` function
is `0.4` for the mean zscore and `0.6` for the standard deviation of the zscore.
Higher weights for the standard deviation of the zscores lead to a greater
emphasis on genes with highly stable correlation patterns.
```{r, select genes, message = FALSE}
sce <- select_genes(
    sce,
    max_n_genes = 2000,
    score_weights = c(0.4, 0.6)
)
# Inspect the anglemania genes
anglemania_genes <- get_anglemania_genes(sce)
head(anglemania_genes)
length(anglemania_genes)
```


# Simpson's paradox correction

anglemania ranks gene pairs by how *consistent* their correlation is across
batches. That works well when every batch contains the same mixture of cell
types, but real datasets rarely do. A gene pair that is correlated only within
T cells will show a correlation proportional to each batch's T cell fraction --
so a pair with a perfectly stable biological relationship looks unstable, and a
pair with no within-cell-type relationship at all can look stable simply because
two batches happen to share a composition. This is Simpson's paradox: the
aggregate correlation is driven by the mixture, not by the biology.

The correction, enabled with `use_simpsons = TRUE`, groups cells into
microclusters within each batch (PCA, then an SNN graph, then Leiden) and
computes the gene-gene correlations with per-cell weights `1 / |cluster|`. Every
microcluster then contributes equally to the estimate regardless of how many
cells it holds, which removes the composition term.

The simulated data used above has balanced composition by construction, so it
cannot show the effect. We first make the batches unbalanced -- each batch is
dominated by a different cell type -- and then compare gene selection with and
without the correction.

```{r, simpsons unbalanced data}
set.seed(1)
batches <- unique(as.character(sce_raw$Batch))
groups <- unique(as.character(sce_raw$Group))

# Each batch keeps all cells of one dominant group and 30% of the others
keep <- unlist(lapply(seq_along(batches), function(i) {
    dominant <- groups[(i - 1) %% length(groups) + 1]
    in_batch <- which(as.character(sce_raw$Batch) == batches[i])
    is_dominant <- as.character(sce_raw$Group)[in_batch] == dominant
    c(
        in_batch[is_dominant],
        sample(in_batch[!is_dominant], round(0.3 * sum(!is_dominant)))
    )
}))

sce_unbalanced <- sce_raw[, keep]
table(sce_unbalanced$Batch, sce_unbalanced$Group)
```

Each batch is now dominated by a different cell type. Running anglemania with
and without the correction:

```{r, run simpsons, message = FALSE}
sce_plain <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000
)

sce_simpsons <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000,
    use_simpsons = TRUE
)

genes_plain <- get_anglemania_genes(sce_plain)
genes_simpsons <- get_anglemania_genes(sce_simpsons)

length(intersect(genes_plain, genes_simpsons))
```

The two gene sets overlap substantially but not completely -- the difference is
the set of genes whose apparent cross-batch stability came from the composition
rather than from the biology. On the SCIB benchmark suite the correction raises
DE gene precision to 0.937-0.983 from a baseline of 0.899-0.901.

## Choosing the number of microclusters

`simpsons_n_clusters` is the parameter worth tuning. It sets the Leiden
resolution to `max(1, simpsons_n_clusters / 20)`. Sweeps across 5-50 clusters
show that 10-20 is optimal: too few and the microclusters are still internally
heterogeneous, so the composition confound survives; too many and each cluster
holds too few cells for a stable correlation estimate. The default is `15`.

`simpsons_min_cells` (default `30`) sets the size below which a microcluster is
merged into its nearest neighbour, and `simpsons_min_cluster_size` (default
`20`) the size below which its cells are dropped from the correlation estimate
entirely. Both matter most for small batches.

## Count splitting on real data

The pipeline above uses the same counts to build the microclusters and to
estimate the correlations within them. That is double-dipping, and on real data
it makes the correction overshoot -- in our PBMC benchmark plain Simpson's
scored ARI 0.180 against a baseline of 0.195. Count splitting
([Neufeld et al.](https://arxiv.org/abs/2307.12985)) fixes this by drawing two
independent Poisson folds from the counts, clustering on the first and
correlating on the second:

```{r, count splitting, eval = FALSE}
sce_simpsons_cs <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000,
    use_simpsons = TRUE,
    simpsons_use_count_split = TRUE,
    simpsons_count_split_prop = 0.5
)
```

On the same PBMC benchmark this recovers the win, reaching ARI 0.205. **If you
are running on real data rather than simulations, turn this on.** It requires
the `countsplit` package.

## Options that are available but not recommended

Two alternatives to the default weighting scheme are exposed as parameters
because they were benchmarked, but neither improved on the default and both
default to `FALSE`:

- `simpsons_use_binary_pca` builds the clustering embedding from the binarised
detection matrix rather than from expression. Cost: 0.4-0.8% DE precision.
- `simpsons_use_density_weights` replaces the discrete cluster-inverse weights
with continuous kNN density weights. Cost: 3-7% DE precision.

Plain Leiden microclusters with cluster-inverse weights remains the best
configuration.

## Using the Simpson's score in gene ranking

Setting `simpsons_weight_in_ranking` above `0` adds a third criterion to
`select_genes()`, alongside the mean and standard deviation of the z-scores. It
scores each gene pair on how strong and how consistent its *within-microcluster*
correlations are across datasets, and the two base weights are rescaled to sum
to `1 - simpsons_weight_in_ranking`. This triggers a post-hoc pass over every
prefiltered gene pair in every microcluster, which is considerably more
expensive than the correction itself, so it is off by default.


# MNN integration
The anglemania genes can now be used for downstream integration algorithms such
as MNN. We compare the integration results using the anglemania genes with the
results using 500 and the standard 2000 HVGs.

## HVGs
### 500 HVGs
```{r, MNN_500 HVGs}
hvg_500 <- sce %>%
    scater::logNormCounts() %>%
    modelGeneVar(block = colData(sce)[[batch_key]]) %>%
    getTopHVGs(n = 500)

barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = hvg_500,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_hvg_500") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_500", name = "umap_MNN_hvg_500")
# k is the number of nearest neighbours to consider when identifying MNNs
```

### 2000 HVGs
```{r, MNN_2000 HVGs}
hvg_2000 <- sce %>%
    scater::logNormCounts() %>%
    modelGeneVar(block = colData(sce)[[batch_key]]) %>%
    getTopHVGs(n = 2000)

barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = hvg_2000,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_hvg_2000") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_2000", name = "umap_MNN_hvg_2000")
```

## anglemania genes
```{r, MNN anglemania, message = FALSE}
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = anglemania_genes,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_anglemania_2000") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(
    sce,
    dimred = "MNN_anglemania_2000",
    name = "umap_MNN_anglemania_2000"
)

sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = anglemania_genes[1:500],
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_anglemania_500") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(
    sce,
    dimred = "MNN_anglemania_500",
    name = "umap_MNN_anglemania_500"
)


```

## Plot
### UMAP embeddings
We can see from the UMAPs that anglemania genes yield the best integration
in terms of clustering by cell type and mixing the batches. The goal of an 
integration and subsequent clustering should be to have low intra cluster
variance and high inter cluster variance. This is at least true for most
downstream scRNA-seq analyses where the goal is to e.g. differentiate between
cell types or cell states and annotate these.
```{r, fig.cap = "UMAPs of MNN integrated data. Comparison of UMAP embeddings of integrated data using anglemania genes, top 500 HVGs and top 2000 HVGs.", fig.width = 12, fig.height = 8, fig.wide = TRUE}
# Use wrap_plots
patchwork::wrap_plots(
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_anglemania_500") +
        ggtitle("MNN integration using 500 anglemania genes, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_anglemania_500") +
        ggtitle("MNN integration using anglemania genes, colored by Group"),
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_500") +
        ggtitle("MNN integration using top 500 HVGs, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_500") +
        ggtitle("MNN integration using top 500 HVGs, colored by Group"),
    ncol = 2
)
patchwork::wrap_plots(
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_anglemania_2000") +
        ggtitle("MNN integration using 2000 anglemania genes, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_anglemania_2000") +
        ggtitle("MNN integration using 2000 anglemania genes, colored by Group"),
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_2000") +
        ggtitle("MNN integration using top 2000 HVGs, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_2000") +
        ggtitle("MNN integration using top 2000 HVGs, colored by Group"),
    ncol = 2
)
```

### Overlap
```{r, fig.cap = "Overlap of selected genes. Additionally, we check the overlap of the anglemania genes with the HVGs. About 211 of the 2000 anglemania genes are also found in the top 500 HVGs, and 822 of the 2000 anglemania genes are also found in the top 2000 HVGs.", fig.width = 10, fig.height = 8, fig.wide = TRUE, message = FALSE, warning = FALSE}
upsetr_df <- fromList(
    list(
        anglemania = anglemania_genes,
        hvg_500 = hvg_500,
        hvg_2000 = hvg_2000
    )
)
upset(upsetr_df, text.scale = 2)
```

# Seurat
Now you can just use the anglemania genes for other integration algorithms.
When using Seurat, the easiest approach is to create an SCE from the counts and metadata
of the SeuratObject, then run anglemania on it and save those genes as the
VariableFeatures of the SeuratObject.
```{r, message = FALSE, warning = FALSE}
se <- CreateSeuratObject(
    counts = counts(sce_raw),
    meta.data = as.data.frame(colData(sce_raw))
)
se
anglemania_genes <- se |>
    as.SingleCellExperiment(assay = "RNA") |>
    anglemania(
        batch_key = "Batch",
        max_n_genes = 2000
    ) |>
    get_anglemania_genes()
```


## Integration
In Seurat v5 you split the layers of an assay by batch and then run the normal
Seurat workflow. To use anglemania genes for integration, you need to assign them 
to the VariableFeatures slot of the SeuratObject. After that, you integrate the layers 
using the anglemania genes as the `features` argument.
```{r, Seurat Integration v5, message = FALSE, warning = FALSE}
# Split by batch
se[["RNA"]] <- split(se[["RNA"]], f = se$Batch)

# Standard preprocessing but use anglemania genes as "VariableFeatures"
se <- NormalizeData(se, verbose = FALSE)
VariableFeatures(se) <- anglemania_genes
se <- se |>
    ScaleData(verbose = FALSE) |>
    RunPCA(verbose = FALSE)

# Integrate
se <- IntegrateLayers(
    object = se,
    method = CCAIntegration,
    orig.reduction = "pca",
    new.reduction = "integrated.cca",
    features = anglemania_genes,
    verbose = FALSE
)
se <- RunUMAP(se, dims = 1:30, reduction = "integrated.cca", verbose = FALSE)
se
```





## Plot
```{r, fig.cap = "UMAPs of Seurat integrated data. Here we show that we can use the anglemania genes for integration of a SeuratObject.", fig.wide = TRUE, fig.width = 10}
patchwork::wrap_plots(
    DimPlot(se, reduction = "umap", group.by = "Batch") +
        ggtitle("Seurat integration using\nanglemania genes\ncolored by Batch"),
    DimPlot(se, reduction = "umap", group.by = "Group") +
        ggtitle("Seurat integration using\nanglemania genes\ncolored by Group"),
    ncol = 2
)
```



# Showcase underlying functions
## Normal anglemania workflow
```{r, message = FALSE}
sce_raw <- sce_example()
sce <- sce_raw
batch_key <- "batch"
sce <- anglemania(sce, batch_key = batch_key, verbose = FALSE)
```

`anglemania` is run on the SCE object and it basically calls three functions: 

- `factorise`: 
    - creates a permutation of the input matrix whose correlation matrix
    is used to create a null distribution for each batch.
    - computes the cosine similarity (or spearman coefficient) between gene
    expression vector pairs matrix for both the original and permuted matrices
    - computes the zscore of the relationship between the gene pairs taking 
    the mean and standard deviation of the null distribution
    - it does this for every batch in the dataset!
- `get_list_stats`
    - computes the mean and standard deviation of the zscores across the 
    matrices from the different batches. This creates two important matrices:
    the mean zscore matrix `mean_zscore` and the signal-to-noise ratio matrix
    `sn_zscore`. These are stored in the metadata of the SCE object.
- `select_genes`
    - filters the gene pairs by the `mean_zscore` and `sn_zscore` matrices
    (SN ratio, i.e. the mean divided by the standard deviation).

## factorise
```{r, message = FALSE}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
counts_by_batch <- lapply(barcodes_by_batch, function(x) {
    counts(sce[, x]) %>% sparse_to_fbm()
})
counts_by_batch[[1]][1:10, 1:6]
# we are working on FBMs (file-backed matrices
# implemented in the bigstatsr package)
class(counts_by_batch[[1]])

# factorise produces the correlation matrices transformed to z-scores
factorised <- lapply(counts_by_batch, factorise)
factorised[[1]][1:10, 1:6]
```

## get_list_stats
The "list stats" are computed by `get_list_stats` and take the z-score 
transformed correlation matrices from `factorise` as input.
The outputs are the mean zscore matrix `mean_zscore` and the 
signal-to-noise ratio matrix `sn_zscore`. These are stored in the metadata of
the SCE object. 

```{r, message = FALSE}
matrix_list <- metadata(sce)$anglemania$matrix_list
weights <- setNames(
    metadata(sce)$anglemania$params$dataset_weights$weight,
    metadata(sce)$anglemania$params$dataset_weights$anglemania_batch
)
list_stats <- get_list_stats(
    matrix_list = matrix_list,
    weights = weights,
    verbose = FALSE
)
names(list_stats)
class(list_stats)
list_stats$mean_zscore[1:10, 1:6]
list_stats$sn_zscore[1:10, 1:6]

# Or we can access them directly from the SCE object
# after running anglemania
metadata(sce)$anglemania$list_stats$mean_zscore[1:10, 1:6]
metadata(sce)$anglemania$list_stats$sn_zscore[1:10, 1:6]
```

## select_genes
- under the hood, `anglemania` calls `select_genes`
- we can use `select_genes` to change the number of selected genes without
having to run anglemania again
```{r, message = FALSE}
previous_genes <- get_anglemania_genes(sce)
sce <- select_genes(
    sce,
    max_n_genes = 2000,
    verbose = FALSE
)
# Inspect the anglemania genes
new_genes <- get_anglemania_genes(sce)

length(previous_genes)
length(new_genes)
```


# sessionInfo
```{r}
sessionInfo()
```
