---
title: "anglemania_tutorial_SCE"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{anglemania tutorial SCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
.libPaths(.Library)
```

# Introduction
anglemania is a feature selection package that extracts genes from multi-batch scRNA-seq 
experiments for downstream dataset integration.
The goal is to select genes that carry high biological information and low 
technical noise between the batches. Those genes are extracted from gene pairs 
that have an invariant and extremely narrow or wide angle between their expression vectors.
Conventionally, highly-variable genes (HVGs) or sometimes all genes are used
for integration tasks (https://www.nature.com/articles/s41592-021-01336-8).
While HVGs are a great and easy way to reduce the noise and
dimensionality of the data, they are not optimal for integration tasks. HVGs are
sensitive to batch effects because the variance is a function of both the technical
and biological variance.
anglemania improves conventional usage of HVGs for integration tasks, 
especially when the transcriptional difference between cell types or cell states 
is subtle (showcased here with de.facLoc set to 0.1 (= mild differences between "Groups")).
The package can be used on top of SingleCellExperiment or Seurat objects.

Under the hood, anglemania works with file-backed big matrices (FBMs) from the bigstatsr package 
(https://github.com/privefl/bigstatsr/) for fast and memory efficient computation.

```{r, load libraries}
suppressPackageStartupMessages({
  library(anglemania)
  library(dplyr)
  library(Seurat)
  library(splatter)
  library(SingleCellExperiment)
  library(scater)
  library(scran)
  library(bluster)
  library(RcppPlanc)
  library(rliger)
  library(batchelor)
})
```

# Create simulated data with mild batch effects and mild differences between the groups/cell types
```{r, create simulated data}
batch.facLoc <- 0.1
de.facLoc <- 0.1
nBatches <- 4
nGroups <- 3
nGenes <- 5000
groupCells <- 300

sce <- splatSimulate(
    batchCells = rep(groupCells * nGroups, nBatches),
    batch.facLoc = batch.facLoc,
    group.prob = rep(1/nGroups, nGroups),
    nGenes = nGenes,
    batch.facScale = 0.1,
    method = "groups",
    verbose = FALSE,
    out.prob    = 0.001,
    de.prob     = 0.1, # mild
    de.facLoc   = de.facLoc,
    de.facScale = 0.1,
    bcv.common  = 0.1,
    seed = 42
)
sce
```

# unintegrated data
```{r}
sce_unintegrated <- sce
# Normalization.
sce_unintegrated <- logNormCounts(sce_unintegrated)

# Feature selection.
dec <- modelGeneVar(sce_unintegrated)
hvg <- getTopHVGs(dec, prop = 0.1)

# PCA.
set.seed(1234)
sce_unintegrated <- scater::runPCA(sce_unintegrated, ncomponents = 50, subset_row = hvg)

# Clustering.
colLabels(sce_unintegrated) <- clusterCells(sce_unintegrated,
  use.dimred = "PCA",
  BLUSPARAM = NNGraphParam(cluster.fun = "louvain")
)

# Visualization.
sce_unintegrated <- scater::runUMAP(sce_unintegrated, dimred = "PCA")
plotUMAP(sce_unintegrated, colour_by = "Batch")
plotUMAP(sce_unintegrated, colour_by = "Group")
```

# run anglemania
## create anglemania object
* We create an anglemania object to have a standardized input to the anglemania function
  * It can be created from:
    * An SingleCellExperiment (SCE) object
    * Seurat object
    * a list of SCE and/or Seurat objects
* pre-requisites are that you:
  * have a column in the metadata (colData for SCE) that specifies the batch/sample a cell belongs to
  * or if you pass a list of SCE/Seurat objects, just provide this as the input and
   don't specify a batch key
```{r, create anglemania object}
head(colData(sce))
batch_key <- "Batch"
angl <- create_anglemania_object(
  sce,
  batch_key = batch_key
)
angl
```

## run anglemania function
* the anglemania function has five arguments
  * the input anglemania object
  * method: either cosine, spearman or diem
    * this is the method by which the relationship of the gene pairs is measured. Default is cosine, 
      which is the cosine similarity between the expression vectors of the gene pairs.
  * zscore_mean_threshold: We compute a mean of the zscore of the relationship between a gene pair,
    and then we set a minimal cutoff for the (absolute) mean zscore. A cutoff of 2 means that the
    filtered gene pairs have a relationship, e.g. cosine similarity, that is 2 standard deviations 
    away from the mean of all cosine similarities from this dataset. A higher value means a more 
    stringent cutoff.
  * zscore_sn_threshold: The SN or signal-to-noise ratio measures the invariance of the relationship
    of the relationship between the gene pair. A high SN ratio means that the relationship is constant
    over multiple batches.
  * max_n_genes: you can specify a maximum number of extracted genes. They are sorted by decreasing 
    mean zscore
```{r, run anglemania}
angl <- anglemania(angl,
  zscore_mean_threshold = 2,
  zscore_sn_threshold = 2,
  max_n_genes = 2000 # optionally define a max number of genes. default is 2000
)
# If you think the number of selected genes is too high or low you can adjust the thresholds:
angl <- select_genes(angl,
                     zscore_mean_threshold = 2.5,
                     zscore_sn_threshold = 2.5)
# Inspect the anglemania genes
integration_genes <- get_anglemania_genes(angl)
head(integration_genes)
length(integration_genes)
```

* We use LIGER https://github.com/welch-lab/liger, https://www.nature.com/articles/s41596-020-0391-8 
  and fastMNN https://rdrr.io/github/LTLA/batchelor/, https://www.nature.com/articles/nbt.4091
  and compare the results when we use HVGs and anglemania genes as the input

# Liger integration
## HVGs
### low number of HVGs
```{r, echo=FALSE, results="hide"}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[,x])
sce_lig <- rliger::createLiger(sce_list, "rna")
sce_lig <- sce_lig %>%
  rliger::normalize() %>%
  rliger::selectGenes(nGenes = 320) %>%
  # because we chose 319 genes with anglemania we only keep 320 HVGs for better comparability
  rliger::scaleNotCenter()
sce_lig <- rliger::runIntegration(sce_lig, k = 20)
sce_lig
```

```{r, echo=FALSE, results="hide"}
sce_lig <- runUMAP(sce_lig)
sce_lig
umap_rliger <- rliger::dimRed(sce_lig, "UMAP")
rownames(umap_rliger) <- sapply(rownames(umap_rliger), function(x) stringr::str_split(x, "_")[[1]][2])
reducedDim(sce, "umap_liger_hvg_320") <- umap_rliger
```

### 2000 HVGs 
```{r, echo=FALSE, results="hide"}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[,x])
sce_lig <- rliger::createLiger(sce_list, "rna")
sce_lig <- sce_lig %>%
  rliger::normalize() %>%
  rliger::selectGenes(nGenes = 2000) %>%
  rliger::scaleNotCenter()
sce_lig <- rliger::runIntegration(sce_lig, k = 20)
sce_lig
```

```{r, echo=FALSE, results="hide"}
sce_lig <- runUMAP(sce_lig)
sce_lig
umap_rliger <- rliger::dimRed(sce_lig, "UMAP")
rownames(umap_rliger) <- sapply(rownames(umap_rliger), function(x) stringr::str_split(x, "_")[[1]][2])
reducedDim(sce, "umap_liger_hvg_2000") <- umap_rliger
```


## anglemania
```{r, echo=FALSE, results="hide"}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[,x])
sce_lig <- rliger::createLiger(sce_list, "rna")
sce_lig <- rliger::normalize(sce_lig)
varFeatures(sce_lig) <- integration_genes # just act like the anglemania genes are the "variable features"
sce_lig <- rliger::scaleNotCenter(sce_lig)
sce_lig <- rliger::runIntegration(sce_lig, k = 20)
sce_lig
```

```{r, echo=FALSE, results="hide"}
sce_lig <- runUMAP(sce_lig)
sce_lig
umap_rliger <- rliger::dimRed(sce_lig, "UMAP")
rownames(umap_rliger) <- sapply(rownames(umap_rliger), function(x) stringr::str_split(x, "_")[[1]][2])
reducedDim(sce, "umap_liger_anglemania") <- umap_rliger
```

## Plot
```{r}
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_liger_hvg_320") +
  ggtitle("Integration using top 320 HVGs, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_liger_hvg_320") +
  ggtitle("Integration using top 320 HVGs, colored by Group")
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_liger_hvg_2000") +
  ggtitle("Integration using top 2000 HVGs, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_liger_hvg_2000") +
  ggtitle("Integration using top 2000 HVGs, colored by Group")
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_liger_anglemania") +
  ggtitle("Integration using anglemania genes, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_liger_anglemania") +
  ggtitle("Integration using anglemania genes, colored by Group")
```

# MNN integration
## HVGs
### 320 HVGs
```{r, echo=FALSE, results="hide"}
hvg <- sce %>%
  scater::logNormCounts() %>%
  modelGeneVar(block = colData(sce)[[batch_key]]) %>%
  getTopHVGs(n = 320)
  
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
  scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(sce_mnn, subset.row = hvg, k = 20, batch = factor(colData(sce_mnn)[[batch_key]]), d = 50)
reducedDim(sce, "MNN_hvg_320") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_320", name = "umap_MNN_hvg_320")
# k is the number of nearest neighbours to consider when identifying MNNs
```

### 2000 HVGs
```{r, echo=FALSE, results="hide"}
hvg <- sce %>%
  scater::logNormCounts() %>%
  modelGeneVar(block = colData(sce)[[batch_key]]) %>%
  getTopHVGs(n = 2000)
  
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
  scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(sce_mnn, subset.row = hvg, k = 20, batch = factor(colData(sce_mnn)[[batch_key]]), d = 50)
reducedDim(sce, "MNN_hvg_2000") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_2000", name = "umap_MNN_hvg_2000")
```

## anglemania
```{r, echo=FALSE, results="hide"}
sce_mnn <- sce %>%
  scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(sce_mnn, subset.row = integration_genes, k = 20, batch = factor(colData(sce_mnn)[[batch_key]]), d = 50)
reducedDim(sce, "MNN_anglemania") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_anglemania", name = "umap_MNN_anglemania")
```

## Plot
```{r}
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_320") +
  ggtitle("MNN integration using top 320 HVGs, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_320") +
  ggtitle("MNN integration using top 320 HVGs, colored by Group")
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_2000") +
  ggtitle("MNN integration using top 2000 HVGs, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_2000") +
  ggtitle("MNN integration using top 2000 HVGs, colored by Group")
plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_anglemania") +
  ggtitle("MNN integration using anglemania genes, colored by Batch")
plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_anglemania") +
  ggtitle("MNN integration using anglemania genes, colored by Group")
```


```{r}
sessionInfo()
```
