---
title: "anglemania Tutorial"
author:
- name: Aaron Kollotzek
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
  email: aaron.kollotzek@mdc-berlin.de
- name: Vedran Franke
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
- name: Artem Baranovskii
  affiliation:
  - Helmholtz Munich
- name: Altuna Akalin
  affiliation:
  - &MDC Max Delbrück Center for Molecular Medicine in the Helmholtz Association, Berlin, Germany
  - &BIMSB Berlin Institute for Medical Systems Biology, Berlin, Germany
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{anglemania tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
anglemania is a feature selection package that extracts genes from multi-batch scRNA-seq 
experiments for downstream dataset integration.
The goal is to select genes that carry high biological information and low 
technical noise between the batches. Those genes are extracted from gene pairs 
that have an invariant and extremely narrow or wide angle between their expression vectors.
Conventionally, highly-variable genes (HVGs) or sometimes all genes are used
for integration tasks (https://www.nature.com/articles/s41592-021-01336-8).
While HVGs are a great and easy way to reduce the noise and
dimensionality of the data, we hypothesize that there are better ways to select genes
specifically for integration tasks. HVGs are sensitive to batch effects because the
variance is a function of both the technical and biological variance.
anglemania improves conventional usage of HVGs for integration tasks, 
especially when the transcriptional difference between cell types or cell states 
is subtle (showcased here using simulated data generated using `r Biocpkg("splatter")`
with `de.facLoc` and `de.facScale` set to 0.1, which results in mild differences between 
"Groups").
The package can be used on top of `r Biocpkg("SingleCellExperiment")` or
`r CRANpkg("Seurat")` objects.

Under the hood, anglemania works with file-backed big matrices (FBMs) from
the `r CRANpkg("bigstatsr")` package for fast and memory 
efficient computation.

```{r, load libraries}
suppressPackageStartupMessages({
    library(anglemania)
    library(dplyr)
    library(Seurat)
    options(Seurat.object.assay.version = "v5")
    library(splatter)
    library(SingleCellExperiment)
    library(scater)
    library(scran)
    library(bluster)
    library(batchelor)
    library(UpSetR)
})
```

# Simulation
We simulate a scRNA-seq dataset using Splatter with 4 batches and 3 cell types with subtle differences 
between cell types and rather big batch effects. 
```{r, create simulated data}
batch.facLoc <- 0.3
de.facLoc <- 0.1
nBatches <- 4
nGroups <- 3
nGenes <- 5000
groupCells <- 300

sce_raw <- splatSimulate(
    batchCells = rep(groupCells * nGroups, nBatches),
    batch.facLoc = batch.facLoc,
    group.prob = rep(1 / nGroups, nGroups),
    nGenes = nGenes,
    batch.facScale = 0.1,
    method = "groups",
    verbose = FALSE,
    out.prob = 0.001,
    de.prob = 0.1, # mild
    de.facLoc = de.facLoc,
    de.facScale = 0.1,
    bcv.common = 0.1,
    seed = 42
)
sce <- sce_raw
assays(sce)
```

# Unintegrated data
Here we perform a standard workflow on the unintegrated data. 
When we perform clustering on the unintegrated data and visualize it in a UMAP, 
we can see that the clusters are driven by batch effects rather than cell types.
```{r, fig.cap = "UMAPs of unintegrated data, colored by Batch and Group. The clusters are driven by batch effects.", fig.wide = TRUE, fig.width = 10}
sce_unintegrated <- sce
# Normalization.
sce_unintegrated <- logNormCounts(sce_unintegrated)

# Feature selection.
dec <- modelGeneVar(sce_unintegrated)
hvg <- getTopHVGs(dec, prop = 0.1)

# PCA.
set.seed(1234)
sce_unintegrated <- scater::runPCA(
    sce_unintegrated,
    ncomponents = 50,
    subset_row = hvg
)

# Clustering.
colLabels(sce_unintegrated) <- clusterCells(sce_unintegrated,
    use.dimred = "PCA",
    BLUSPARAM = NNGraphParam(cluster.fun = "louvain")
)

# Visualization.
sce_unintegrated <- scater::runUMAP(sce_unintegrated, dimred = "PCA")
patchwork::wrap_plots(
    plotUMAP(sce_unintegrated, colour_by = "Batch") +
        ggtitle("Unintegrated data, colored by Batch"),
    plotUMAP(sce_unintegrated, colour_by = "Group") +
        ggtitle("Unintegrated data, colored by Group")
)
```



# anglemania
anglemania works on a `r Biocpkg("SingleCellExperiment")` object. 
The function has a few important arguments: 
- `batch_key`: the column in the metadata of the SingleCellExperiment 
object that indicates which batch the cells belong to. This is required to 
distinguish between batches, because we compute the angle between gene pairs 
for each batch.
- `method`: either cosine or spearman
    - this is the method by which the relationship of the gene pairs is measured. 
Default is cosine, which is the cosine similarity between the expression 
vectors of the gene pairs.
- `max_n_genes`: you can specify a maximum number of extracted genes. They are 
sorted by a weighted sum of the ranked mean and standard deviation of the zscores
of the gene pair correlations.

```{r, run anglemania, message = FALSE}
head(colData(sce))
batch_key <- "Batch"
sce <- anglemania(
    sce,
    batch_key = batch_key,
    max_n_genes = 2000
)
```

## extract the anglemania genes from the SCE object
```{r}
anglemania_genes <- get_anglemania_genes(sce)
head(anglemania_genes)
length(anglemania_genes)
```

## select_genes 
Once anglemania was run on the SCE, you can adjust the number of genes you want
to select using the `select_genes()` function. It has an argument score_weights,
which let's you decide how much weight to set to the mean and standard
deviation of the zscores. The default which is used in the `anglemania` function
is `0.4` for the mean zscore and `0.6` for the standard deviation of the zscore.
Higher weights for the standard deviation of the zscores lead to a greater
emphasis on genes with highly stable correlation patterns.
```{r, select genes, message = FALSE}
sce <- select_genes(
    sce,
    max_n_genes = 2000,
    score_weights = c(0.4, 0.6)
)
# Inspect the anglemania genes
anglemania_genes <- get_anglemania_genes(sce)
head(anglemania_genes)
length(anglemania_genes)
```


# Simpson's paradox correction

anglemania ranks gene pairs by how *consistent* their correlation is across
batches. That works well when every batch contains the same mixture of cell
types, but real datasets rarely do. A gene pair that is correlated only within
T cells will show a correlation proportional to each batch's T cell fraction --
so a pair with a perfectly stable biological relationship looks unstable, and a
pair with no within-cell-type relationship at all can look stable simply because
two batches happen to share a composition. This is Simpson's paradox: the
aggregate correlation is driven by the mixture, not by the biology.

The correction, enabled with `use_simpsons = TRUE`, groups cells into clusters
within each batch (PCA, then an SNN graph, then Leiden) and computes the
gene-gene correlations with per-cell weights `1 / |cluster|`. Every cluster
then contributes equally to the estimate regardless of how many cells it holds,
which removes the composition term.

## Seeing the confound directly

Before using the correction it is worth watching it work on the quantity it is
meant to fix. We simulate four batches whose share of one cell type falls from
70% to 15%, take the genes most strongly up-regulated in that cell type, and
measure how much each pair's correlation moves across batches -- with and
without the weights. That cross-batch spread is exactly what anglemania's
standard-deviation criterion penalises.

```{r, simpsons mechanism, message = FALSE}
sim_full <- splatSimulate(
    batchCells = rep(1000 * 3, 4),
    batch.facLoc = 0.3, batch.facScale = 0.1,
    group.prob = rep(1 / 3, 3),
    nGenes = 5000, method = "groups", verbose = FALSE,
    out.prob = 0.001, de.prob = 0.1,
    de.facLoc = 0.3, de.facScale = 0.1, bcv.common = 0.1, seed = 42
)

# Batches with a decreasing share of Group1
set.seed(1)
share <- c(0.70, 0.50, 0.30, 0.15)
grps <- unique(as.character(sim_full$Group))
btch <- unique(as.character(sim_full$Batch))
keep <- unlist(lapply(seq_along(btch), function(i) {
    in_b <- which(as.character(sim_full$Batch) == btch[i])
    g <- as.character(sim_full$Group)[in_b]
    c(
        sample(in_b[g == grps[1]], round(share[i] * 1000)),
        sample(in_b[g == grps[2]], round((1 - share[i]) * 500)),
        sample(in_b[g == grps[3]], round((1 - share[i]) * 500))
    )
}))
sim_demo <- scater::logNormCounts(sim_full[, keep])
table(sim_demo$Batch, sim_demo$Group)
```

Weighted Pearson correlation, the quantity `use_simpsons = TRUE` substitutes for
the ordinary one:

```{r, weighted correlation helper}
wcor <- function(mat, w) {
    w <- w / sum(w)
    centred <- mat - as.numeric(mat %*% w)
    covm <- (centred * rep(w, each = nrow(mat))) %*% t(centred)
    sds <- sqrt(diag(covm))
    covm / outer(sds, sds)
}
```

For each batch we cluster, derive the weights, and record every pair's
correlation both ways:

```{r, compare stability, message = FALSE}
de_genes_g1 <- rownames(sim_demo)[
    order(rowData(sim_demo)$DEFacGroup1, decreasing = TRUE)[1:40]
]
pairs <- t(combn(length(de_genes_g1), 2))
unw <- wtd <- matrix(NA_real_, nrow(pairs), length(btch))

for (i in seq_along(btch)) {
    cells <- colnames(sim_demo)[sim_demo$Batch == btch[i]]
    expr <- as.matrix(logcounts(sim_demo)[de_genes_g1, cells])

    clusters <- create_microclusters_sparse(
        counts(sim_demo)[, cells],
        n_pcs = 20, min_cells = 15, verbose = FALSE
    )
    w <- compute_simpsons_weights(clusters$assignments, min_cluster_size = 10)

    cu <- cor(t(expr))
    cw <- wcor(expr, w[cells])
    unw[, i] <- cu[cbind(pairs[, 1], pairs[, 2])]
    wtd[, i] <- cw[cbind(pairs[, 1], pairs[, 2])]
}

data.frame(
    correlation = c("unweighted", "cluster-weighted"),
    mean_cross_batch_sd = c(mean(apply(unw, 1, sd)), mean(apply(wtd, 1, sd)))
)
```

The cluster-weighted correlations move roughly 17% less across batches
(mean SD 0.035 against 0.042). These are gene pairs whose biology is identical
in every batch; all that changed is the cell type mixture, and the weighting
absorbs most of that. Since anglemania rewards low cross-batch spread, pairs
like these rank higher once the confound is removed.

**This effect needs cells.** Repeating the same comparison at 300 cells per
group instead of 1000 makes it vanish into noise (weighted SD 0.056 against
0.054 -- marginally *worse*). Each cluster's contribution is rescaled to
`1 / |cluster|`, so a cluster holding few cells is promoted to full weight
while still being a noisy estimate. Batches with small or highly skewed
populations can therefore come out worse rather than better.

## How the correction behaves as imbalance grows

The demonstration above used one fixed level of imbalance. The more useful
question is how the correction behaves as batches become progressively more
disproportionate. We sweep that directly, holding the number of cells per batch
**fixed at 900** so that only the proportions change -- otherwise increasing
imbalance would also mean shrinking the data, and the two effects could not be
told apart.

At a dominant share of 1/3 the three cell types are balanced and there is no
confound to remove; at 0.9 each batch is almost a pure population.

```{r, imbalance sweep, message = FALSE}
sim_all <- scater::logNormCounts(sim_full)
cells_per_batch <- 900

sweep_res <- do.call(rbind, lapply(c(1 / 3, 0.5, 0.7, 0.9), function(d) {
    set.seed(1)
    unw <- wtd <- matrix(NA_real_, nrow(pairs), length(btch))
    smallest <- Inf

    for (i in seq_along(btch)) {
        in_b <- which(as.character(sim_all$Batch) == btch[i])
        g <- as.character(sim_all$Group)[in_b]
        dom <- grps[(i - 1) %% length(grps) + 1]
        oth <- setdiff(grps, dom)

        idx <- c(
            sample(in_b[g == dom], round(d * cells_per_batch)),
            sample(in_b[g == oth[1]], round((1 - d) * cells_per_batch / 2)),
            sample(in_b[g == oth[2]], round((1 - d) * cells_per_batch / 2))
        )
        cells <- colnames(sim_all)[idx]
        expr <- as.matrix(logcounts(sim_all)[de_genes_g1, cells])

        clusters <- create_microclusters_sparse(
            counts(sim_all)[, cells], n_pcs = 20, min_cells = 15,
            verbose = FALSE
        )
        w <- compute_simpsons_weights(clusters$assignments,
                                      min_cluster_size = 10)
        smallest <- min(smallest, min(as.numeric(clusters$sizes)))

        cu <- cor(t(expr))
        cw <- wcor(expr, w[cells])
        unw[, i] <- cu[cbind(pairs[, 1], pairs[, 2])]
        wtd[, i] <- cw[cbind(pairs[, 1], pairs[, 2])]
    }

    data.frame(
        dominant_share = round(d, 2),
        unweighted_sd = mean(apply(unw, 1, sd)),
        weighted_sd = mean(apply(wtd, 1, sd)),
        smallest_cluster = smallest
    )
}))

sweep_res$change_pct <- round(
    100 * (sweep_res$weighted_sd - sweep_res$unweighted_sd) /
        sweep_res$unweighted_sd, 1
)
sweep_res
```

The behaviour is not monotone, and it is not uniformly reliable either:

| Dominant share | Unweighted SD | Weighted SD | Change | Smallest cluster |
|---|---|---|---|---|
| 0.33 (balanced) | 0.0334 | 0.0334 | −0.1% | 296 |
| 0.50 | 0.0388 | 0.0357 | **−8.1%** | 223 |
| 0.70 | 0.0429 | 0.0392 | −8.6% | 132 |
| 0.90 | 0.0380 | 0.0478 | **+25.8%** | 45 |

Repeating the sweep under three sampling seeds shows which of these rows can be
trusted. Reading the change column for seeds 1, 2 and 3:

| Dominant share | seed 1 | seed 2 | seed 3 | verdict |
|---|---|---|---|---|
| 0.33 | −0.1% | 0.0% | 0.0% | no effect, completely reproducible |
| 0.50 | −8.1% | −11.7% | −7.8% | consistent help |
| 0.70 | −8.6% | **+8.9%** | −0.8% | **unstable -- the sign flips** |
| 0.90 | +25.8% | +33.9% | +32.2% | consistent harm |

Four things to take from this:

**Balanced batches: the correction does nothing.** To three decimal places, and
identically under every seed. That is the right behaviour -- there is no
composition confound to remove. It also means switching the correction on costs
nothing but compute when your batches happen to be well matched.

**Moderate imbalance is where it works.** Around a 0.5 dominant share it
reliably cuts cross-batch spread by roughly 8-12%. Note the unweighted spread
climbing from 0.0334 to 0.0388 as the confound appears: that increase is exactly
the artefact anglemania would otherwise mistake for unstable biology.

**Strong imbalance is a coin flip.** At a 0.7 dominant share the result swings
between −8.6% and +8.9% depending only on which cells were sampled. The last
column explains why: the smallest cluster in that setting ranged from 82 to 132
cells across seeds, and whether the run helps or hurts tracks whether the
clustering happened to produce a small cluster. Do not rely on the correction
here without checking it on your own data.

**Extreme imbalance actively hurts**, by 26-34%. At a 0.9 dominant share the
minority clusters hold about 45 cells, and cluster-inverse weighting promotes
those 45 cells to the same total influence as the several hundred cells of the
dominant type. The estimate ends up dominated by the noisiest part of the data.

### The same sweep end to end

The comparison above isolates the correlation step. Running the full gene
selection pipeline over the same kind of sweep -- SCIB Sim2 generative
parameters, 4 batches, 4 cell types, **5000 cells per batch throughout**, only
the proportions changing -- and scoring the 1000 selected genes against the
simulator's true DE genes:

| Dominant share | Minority share | Minority cells per batch | Baseline | Simpson's | Gain |
|---|---|---|---|---|---|
| 0.25 (balanced) | 0.25 | 1250 | 1.000 | 1.000 | +0.000 |
| 0.40 | 0.20 | 1000 | 0.990 | 1.000 | +0.010 |
| 0.55 | 0.15 | 750 | 0.948 | 0.993 | +0.045 |
| 0.70 | 0.10 | 500 | 0.901 | 0.934 | +0.033 |
| 0.85 | 0.05 | 250 | 0.836 | 0.897 | **+0.061** |

Two things stand out. First, the **baseline** degrades steadily with imbalance,
1.000 down to 0.836 -- confirming that composition confounding, and nothing
else, is what costs anglemania precision on these data. When the batches are
balanced anglemania is perfect. Second, the correction helps at every level and
never hurts, with the gain broadly growing as the confound does.

That looks like it contradicts the +25.8% harm at a 0.9 dominant share above,
and the difference is the point. Here each batch holds 5000 cells, so even a 5%
minority is 250 cells -- ample. In the correlation sweep each batch held 900
cells, so a 10% minority was around 45 cells, and the weighting promoted those
45 noisy cells to full influence.

So the two sweeps do not disagree; they vary different things. Skew in
*proportion* is what the correction fixes, and it fixes more of it the worse the
skew gets. Skew that drives the minority *count* down towards a few dozen cells
is what breaks it.

The rule of thumb is that the correction needs minority populations that are
*small in proportion but not small in number*. The proportion is what it
corrects; the absolute count is what determines whether the corrected estimate
is any good. Check the realised cluster sizes before trusting it on a badly
skewed dataset:

```{r, check cluster sizes, eval = FALSE}
clusters <- create_microclusters_sparse(counts(sce)[, batch_cells],
                                        verbose = FALSE)
sort(clusters$sizes)
```


## Using the correction

```{r, run simpsons, message = FALSE}
sce_unbalanced <- sim_demo

sce_plain <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000,
    min_cells_per_gene = 10
)

sce_simpsons <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000,
    min_cells_per_gene = 10,
    use_simpsons = TRUE
)

length(intersect(
    get_anglemania_genes(sce_plain),
    get_anglemania_genes(sce_simpsons)
))
```

The two gene sets overlap substantially but not completely. This simulation is
too small to separate the two on DE-gene recovery -- for that, see the
benchmark below.

## Benchmark results

On the two SCIB integration-benchmark simulations, selecting 1000 genes and
scoring against the simulator's ground-truth DE genes
(`Scripts/Reports/test_scib_simulations.Rmd`):

| Simulation | Baseline (permutation) | Simpson's | HVG |
|---|---|---|---|
| Sim1 (6 batches, 7 types) | 0.899 | **0.937** | 0.792 |
| Sim2 (4 batches, 4 types) | 0.901 | **0.983** | 0.696 |

Both simulations have unequal, batch-specific cell type proportions, and both
have thousands of cells per batch.

## Cluster granularity: leave `simpsons_n_clusters` alone

By default (`simpsons_n_clusters = NULL`) the clustering is whatever Leiden
finds at resolution 1, which on these data resolves whole cell types -- a
handful of clusters, not dozens. That is deliberate, and it is the setting all
the benchmark numbers above were produced with.

Supplying an integer instead searches for the Leiden resolution giving roughly
that many clusters. It works -- but it makes results substantially worse. On
SCIB Sim2:

| Clustering | DE precision |
|---|---|
| **`NULL`** (natural communities, 4 clusters) | **0.983** |
| no correction at all | 0.903 |
| `simpsons_n_clusters = 15` | 0.812 |
| `simpsons_n_clusters = 30` | 0.800 |
| `simpsons_n_clusters = 60` | 0.857 |

Forcing genuine microclusters is worse than not correcting at all. The reason is
that the correction works by giving each *cell type* equal weight regardless of
how many cells the batch happens to contain. Split those cell types into finer
groups and you remove the between-cell-type variation that carries the
gene-gene correlation signal, leaving the weighted correlation dominated by
noise.

The name is therefore somewhat misleading: this is cell-type reweighting, not
microclustering. You can check what you actually get:

```{r, inspect clustering, message = FALSE}
b1 <- colnames(sim_demo)[sim_demo$Batch == "Batch1"]
cl <- create_microclusters_sparse(
    counts(sim_demo)[, b1], n_pcs = 20, min_cells = 15, verbose = FALSE
)
cl$n_clusters
```

`simpsons_min_cells` (default `30`) sets the size below which a cluster is
merged into its nearest neighbour, and `simpsons_min_cluster_size` (default
`20`) the size below which its cells are dropped from the correlation estimate.
Given the note above about small clusters, these are the arguments that matter.

## Count splitting on real data

The pipeline above uses the same counts to build the clusters and to estimate
the correlations within them. That is double-dipping, and on real data it makes
the correction overshoot -- in the PBMC benchmark plain Simpson's scored ARI
0.180 against a baseline of 0.195. Count splitting
([Neufeld et al.](https://arxiv.org/abs/2307.12985)) fixes this by drawing two
independent Poisson folds from the counts, clustering on the first and
correlating on the second:

```{r, count splitting, eval = FALSE}
sce_simpsons_cs <- anglemania(
    sce_unbalanced,
    batch_key = "Batch",
    max_n_genes = 2000,
    use_simpsons = TRUE,
    simpsons_use_count_split = TRUE,
    simpsons_count_split_prop = 0.5
)
```

On the same PBMC benchmark this recovers the win, reaching ARI 0.205. **If you
are running on real data rather than simulations, turn this on.** It requires
the `countsplit` package.

## Options that are available but not recommended

Two alternatives to the default weighting scheme are exposed as parameters
because they were benchmarked, but neither improved on it. DE precision on the
SCIB simulations, against 0.937 (Sim1) and 0.983 (Sim2) for the default:

| Option | Sim1 | Sim2 |
|---|---|---|
| `simpsons_use_binary_pca` | 0.929 | 0.979 |
| `simpsons_use_density_weights` | 0.905 | 0.917 |
| both together | 0.902 | 0.899 |

Plain Leiden clusters with cluster-inverse weights remains the best
configuration, and both options default to `FALSE`.

## Using the Simpson's score in gene ranking

Setting `simpsons_weight_in_ranking` above `0` adds a third criterion to
`select_genes()`, alongside the mean and standard deviation of the z-scores. It
scores each gene pair on how strong and how consistent its *within-cluster*
correlations are across datasets, and the two base weights are rescaled to sum
to `1 - simpsons_weight_in_ranking`. This triggers a post-hoc pass over every
prefiltered gene pair in every cluster, which is considerably more expensive
than the correction itself, so it is off by default.


# MNN integration
The anglemania genes can now be used for downstream integration algorithms such
as MNN. We compare the integration results using the anglemania genes with the
results using 500 and the standard 2000 HVGs.

## HVGs
### 500 HVGs
```{r, MNN_500 HVGs}
hvg_500 <- sce %>%
    scater::logNormCounts() %>%
    modelGeneVar(block = colData(sce)[[batch_key]]) %>%
    getTopHVGs(n = 500)

barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = hvg_500,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_hvg_500") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_500", name = "umap_MNN_hvg_500")
# k is the number of nearest neighbours to consider when identifying MNNs
```

### 2000 HVGs
```{r, MNN_2000 HVGs}
hvg_2000 <- sce %>%
    scater::logNormCounts() %>%
    modelGeneVar(block = colData(sce)[[batch_key]]) %>%
    getTopHVGs(n = 2000)

barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
sce_list <- lapply(barcodes_by_batch, function(x) sce[, x])
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = hvg_2000,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_hvg_2000") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(sce, dimred = "MNN_hvg_2000", name = "umap_MNN_hvg_2000")
```

## anglemania genes
```{r, MNN anglemania, message = FALSE}
sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = anglemania_genes,
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_anglemania_2000") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(
    sce,
    dimred = "MNN_anglemania_2000",
    name = "umap_MNN_anglemania_2000"
)

sce_mnn <- sce %>%
    scater::logNormCounts()
sce_mnn <- batchelor::fastMNN(
    sce_mnn,
    subset.row = anglemania_genes[1:500],
    k = 20,
    batch = factor(colData(sce_mnn)[[batch_key]]),
    d = 50
)
reducedDim(sce, "MNN_anglemania_500") <- reducedDim(sce_mnn, "corrected")
sce <- scater::runUMAP(
    sce,
    dimred = "MNN_anglemania_500",
    name = "umap_MNN_anglemania_500"
)


```

## Plot
### UMAP embeddings
We can see from the UMAPs that anglemania genes yield the best integration
in terms of clustering by cell type and mixing the batches. The goal of an 
integration and subsequent clustering should be to have low intra cluster
variance and high inter cluster variance. This is at least true for most
downstream scRNA-seq analyses where the goal is to e.g. differentiate between
cell types or cell states and annotate these.
```{r, fig.cap = "UMAPs of MNN integrated data. Comparison of UMAP embeddings of integrated data using anglemania genes, top 500 HVGs and top 2000 HVGs.", fig.width = 12, fig.height = 8, fig.wide = TRUE}
# Use wrap_plots
patchwork::wrap_plots(
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_anglemania_500") +
        ggtitle("MNN integration using 500 anglemania genes, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_anglemania_500") +
        ggtitle("MNN integration using anglemania genes, colored by Group"),
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_500") +
        ggtitle("MNN integration using top 500 HVGs, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_500") +
        ggtitle("MNN integration using top 500 HVGs, colored by Group"),
    ncol = 2
)
patchwork::wrap_plots(
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_anglemania_2000") +
        ggtitle("MNN integration using 2000 anglemania genes, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_anglemania_2000") +
        ggtitle("MNN integration using 2000 anglemania genes, colored by Group"),
    plotReducedDim(sce, colour_by = "Batch", dimred = "umap_MNN_hvg_2000") +
        ggtitle("MNN integration using top 2000 HVGs, colored by Batch"),
    plotReducedDim(sce, colour_by = "Group", dimred = "umap_MNN_hvg_2000") +
        ggtitle("MNN integration using top 2000 HVGs, colored by Group"),
    ncol = 2
)
```

### Overlap
```{r, fig.cap = "Overlap of selected genes. Additionally, we check the overlap of the anglemania genes with the HVGs. About 211 of the 2000 anglemania genes are also found in the top 500 HVGs, and 822 of the 2000 anglemania genes are also found in the top 2000 HVGs.", fig.width = 10, fig.height = 8, fig.wide = TRUE, message = FALSE, warning = FALSE}
upsetr_df <- fromList(
    list(
        anglemania = anglemania_genes,
        hvg_500 = hvg_500,
        hvg_2000 = hvg_2000
    )
)
upset(upsetr_df, text.scale = 2)
```

# Seurat
Now you can just use the anglemania genes for other integration algorithms.
When using Seurat, the easiest approach is to create an SCE from the counts and metadata
of the SeuratObject, then run anglemania on it and save those genes as the
VariableFeatures of the SeuratObject.
```{r, message = FALSE, warning = FALSE}
se <- CreateSeuratObject(
    counts = counts(sce_raw),
    meta.data = as.data.frame(colData(sce_raw))
)
se
anglemania_genes <- se |>
    as.SingleCellExperiment(assay = "RNA") |>
    anglemania(
        batch_key = "Batch",
        max_n_genes = 2000
    ) |>
    get_anglemania_genes()
```


## Integration
In Seurat v5 you split the layers of an assay by batch and then run the normal
Seurat workflow. To use anglemania genes for integration, you need to assign them 
to the VariableFeatures slot of the SeuratObject. After that, you integrate the layers 
using the anglemania genes as the `features` argument.
```{r, Seurat Integration v5, message = FALSE, warning = FALSE}
# Split by batch
se[["RNA"]] <- split(se[["RNA"]], f = se$Batch)

# Standard preprocessing but use anglemania genes as "VariableFeatures"
se <- NormalizeData(se, verbose = FALSE)
VariableFeatures(se) <- anglemania_genes
se <- se |>
    ScaleData(verbose = FALSE) |>
    RunPCA(verbose = FALSE)

# Integrate
se <- IntegrateLayers(
    object = se,
    method = CCAIntegration,
    orig.reduction = "pca",
    new.reduction = "integrated.cca",
    features = anglemania_genes,
    verbose = FALSE
)
se <- RunUMAP(se, dims = 1:30, reduction = "integrated.cca", verbose = FALSE)
se
```





## Plot
```{r, fig.cap = "UMAPs of Seurat integrated data. Here we show that we can use the anglemania genes for integration of a SeuratObject.", fig.wide = TRUE, fig.width = 10}
patchwork::wrap_plots(
    DimPlot(se, reduction = "umap", group.by = "Batch") +
        ggtitle("Seurat integration using\nanglemania genes\ncolored by Batch"),
    DimPlot(se, reduction = "umap", group.by = "Group") +
        ggtitle("Seurat integration using\nanglemania genes\ncolored by Group"),
    ncol = 2
)
```



# Showcase underlying functions
## Normal anglemania workflow
```{r, message = FALSE}
sce_raw <- sce_example()
sce <- sce_raw
batch_key <- "batch"
sce <- anglemania(sce, batch_key = batch_key, verbose = FALSE)
```

`anglemania` is run on the SCE object and it basically calls three functions: 

- `factorise`: 
    - creates a permutation of the input matrix whose correlation matrix
    is used to create a null distribution for each batch.
    - computes the cosine similarity (or spearman coefficient) between gene
    expression vector pairs matrix for both the original and permuted matrices
    - computes the zscore of the relationship between the gene pairs taking 
    the mean and standard deviation of the null distribution
    - it does this for every batch in the dataset!
- `get_list_stats`
    - computes the mean and standard deviation of the zscores across the 
    matrices from the different batches. This creates two important matrices:
    the mean zscore matrix `mean_zscore` and the signal-to-noise ratio matrix
    `sn_zscore`. These are stored in the metadata of the SCE object.
- `select_genes`
    - filters the gene pairs by the `mean_zscore` and `sn_zscore` matrices
    (SN ratio, i.e. the mean divided by the standard deviation).

## factorise
```{r, message = FALSE}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
counts_by_batch <- lapply(barcodes_by_batch, function(x) {
    counts(sce[, x]) %>% sparse_to_fbm()
})
counts_by_batch[[1]][1:10, 1:6]
# we are working on FBMs (file-backed matrices
# implemented in the bigstatsr package)
class(counts_by_batch[[1]])

# factorise produces the correlation matrices transformed to z-scores
factorised <- lapply(counts_by_batch, factorise)
factorised[[1]][1:10, 1:6]
```

## get_list_stats
The "list stats" are computed by `get_list_stats` and take the z-score 
transformed correlation matrices from `factorise` as input.
The outputs are the mean zscore matrix `mean_zscore` and the 
signal-to-noise ratio matrix `sn_zscore`. These are stored in the metadata of
the SCE object. 

```{r, message = FALSE}
matrix_list <- metadata(sce)$anglemania$matrix_list
weights <- setNames(
    metadata(sce)$anglemania$params$dataset_weights$weight,
    metadata(sce)$anglemania$params$dataset_weights$anglemania_batch
)
list_stats <- get_list_stats(
    matrix_list = matrix_list,
    weights = weights,
    verbose = FALSE
)
names(list_stats)
class(list_stats)
list_stats$mean_zscore[1:10, 1:6]
list_stats$sn_zscore[1:10, 1:6]

# Or we can access them directly from the SCE object
# after running anglemania
metadata(sce)$anglemania$list_stats$mean_zscore[1:10, 1:6]
metadata(sce)$anglemania$list_stats$sn_zscore[1:10, 1:6]
```

## select_genes
- under the hood, `anglemania` calls `select_genes`
- we can use `select_genes` to change the number of selected genes without
having to run anglemania again
```{r, message = FALSE}
previous_genes <- get_anglemania_genes(sce)
sce <- select_genes(
    sce,
    max_n_genes = 2000,
    verbose = FALSE
)
# Inspect the anglemania genes
new_genes <- get_anglemania_genes(sce)

length(previous_genes)
length(new_genes)
```


# sessionInfo
```{r}
sessionInfo()
```
