---
title: "anglemania_tutorial"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{anglemania tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load libraries}
suppressPackageStartupMessages({
  library(anglemania)
  library(dplyr)
  library(Seurat)
})
```

# load data
## unintegrated data
```{r, load data}
load(system.file("extdata", "seurat_splatter_sim.RData", package = "anglemania"))
se
head(se[[]])
se_unintegrated <- se
se_unintegrated <- se_unintegrated %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
DimPlot(se_unintegrated, reduction = "umap", group.by = "Batch")
DimPlot(se_unintegrated, reduction = "umap", group.by = "Group")
```

# run anglemania
## create anglem object
```{r, create anglem object}
head(se[[]])
batch_key <- "Batch"

angl <- create_anglemaniaObject(se,
  batch_key = batch_key,
  min_cells_per_gene = 1
)
angl
```

## run anglemania
```{r, run anglemania}
angl <- anglemania(angl,
  zscore_mean_threshold = 2.5,
  zscore_sn_threshold = 2.5,
  max_n_genes = 3000 # optionally define a max number of genes. default is 2000
)

# Inspect the anglemania genes
integration_genes <- get_anglemania_genes(angl)
head(integration_genes)
length(integration_genes)
```

## integration
* we implemented the easy-to-use integrate_by_features() function from the anglemania package which uses Seurat CCA integration
* you can also just use the anglemania genes for other integration algorithms
```{r, integration option 1}
# if the Seurat FindIntegrationAnchors() function does not work,
# change this to the specified size:
options(future.globals.maxSize = 8000 * 1024^2)
seurat_integrated_angl <- integrate_by_features(se,
  angl,
  process = TRUE
)
seurat_integrated_angl
```


## plot integration
```{r}
Seurat::DimPlot(seurat_integrated_angl, reduction = "umap", group.by = "Batch")
Seurat::DimPlot(seurat_integrated_angl, reduction = "umap", group.by = "Group")
```

# Comparison to using HVGs
## get HVGs
```{r, get HVGs}
se_list <- SplitObject(se, split.by = batch_key)
se_list <- lapply(se_list, NormalizeData)
hvg_features <- Seurat::SelectIntegrationFeatures(se_list, nfeatures = 2000)
```

## integration
```{r, integration option 2}
seurat_integrated_hvg <- integrate_seurat_list(se_list,
  features = hvg_features,
  process = TRUE
)
seurat_integrated_hvg
```

## plot integration
```{r}
Seurat::DimPlot(seurat_integrated_hvg, reduction = "umap", group.by = "Batch")
Seurat::DimPlot(seurat_integrated_hvg, reduction = "umap", group.by = "Group")
```
