---
title: "Running anglemania with Seurat"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{anglemania tutorial Seurat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# .libPaths(.Library)
```

# Introduction
anglemania is a feature selection package that extracts genes from multi-batch scRNA-seq 
experiments for downstream dataset integration.
The goal is to select genes that carry high biological information and low 
technical noise. Those genes are extracted from gene pairs that have an 
invariant and extremely narrow or wide angle between their expression vectors.
It improves conventional usage of highly-variable genes for integration tasks, 
especially when the transcriptional difference between cell types or cell states 
is subtle (showcased here with de.facLoc set to 0.1 (= mild differences between "Groups")).
anglemania can be used on top of SingleCellExperiment or Seurat objects.

```{r, load libraries}
suppressPackageStartupMessages({
  # library(anglemania)
  devtools::load_all()
  library(dplyr)
  library(Seurat)
  library(splatter)
})
```

# Create simulated data
```{r, create simulated data}
batch.facLoc <- 0.3
de.facLoc <- 0.1
nBatches <- 4
nGroups <- 3
nGenes <- 5000
groupCells <- 300

sim <- splatSimulate(
    batchCells = rep(groupCells * nGroups, nBatches),
    batch.facLoc = batch.facLoc,
    group.prob = rep(1/nGroups, nGroups),
    nGenes = nGenes,
    batch.facScale = 0.1,
    method = "groups",
    verbose = FALSE,
    out.prob    = 0.001,
    de.prob     = 0.1, # mild 
    de.facLoc   = de.facLoc,
    de.facScale = 0.1,
    bcv.common  = 0.1,
    seed = 42
)

se <- CreateSeuratObject(counts = counts(sim), meta.data = as.data.frame(colData(sim)))
# rm(sim)
se
Layers(se)
Seurat::Assays(se)
head(se[[]])
```

# unintegrated data
```{r}

se_unintegrated <- se
suppressWarnings({
  se_unintegrated <- se_unintegrated %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30, verbose = FALSE)
})
DimPlot(se_unintegrated, reduction = "umap", group.by = "Batch")
DimPlot(se_unintegrated, reduction = "umap", group.by = "Group")
```

# run anglemania
## create anglemania object
* We create an anglemania object to have a standardized input to the anglemania function
  * It can be created from:
    * An SingleCellExperiment (SCE) object
    * Seurat object
    * a list of SCE and/or Seurat objects
* pre-requisites are that you:
  * have a column in the metadata (colData for SCE) that specifies the batch/sample a cell belongs to
  * or if you pass a list of SCE/Seurat objects, just provide this as the input and
   don't specify a batch key
  
```{r, create anglem object}
head(se[[]])
batch_key <- "Batch"

angl <- create_anglemania_object(se,
  batch_key = batch_key,
  min_cells_per_gene = 1,
  seurat_assay = "RNA"
)
angl
```

## run anglemania function
* the anglemania function has five arguments
  * the input anglemania object
  * method: either cosine, spearman or diem
    * this is the method by which the relationship of the gene pairs is measured. Default is cosine, 
  * zscore_mean_threshold: We compute a mean of the zscore of the relationship between a gene pair,
    and then we set a minimal cutoff for the (absolute) mean zscore. A cutoff of 2 means that the
    filtered gene pairs have a relationship, e.g. cosine similarity, that is 2 standard deviations 
    away from the mean of all cosine similarities from this dataset. A higher value means a more 
    stringent cutoff.
  * zscore_sn_threshold: The SN or signal-to-noise ratio measures the invariance of the relationship
    of the relationship between the gene pair. A high SN ratio means that the relationship is constant
    over multiple batches.
  * max_n_genes: you can specify a maximum number of extracted genes. They are sorted by decreasing 
    mean zscore
```{r, run anglemania}
angl <- anglemania(angl,
  zscore_mean_threshold = 2,
  zscore_sn_threshold = 2,
  max_n_genes = 2000 # optionally define a max number of genes.
)
angl <- select_genes(angl,
                     zscore_mean_threshold = 2.5,
                     zscore_sn_threshold = 2.5)
# Inspect the anglemania genes
integration_genes <- get_anglemania_genes(angl)
head(integration_genes)
length(integration_genes)
```

## integration
* we implemented the easy-to-use integrate_by_features() function from the anglemania package which uses Seurat CCA integration
* you can also just use the anglemania genes for other integration algorithms
```{r, integration option 1}
# if the Seurat FindIntegrationAnchors() function does not work,
# change this to the specified size:
options(future.globals.maxSize = 4000 * 1024^2)
suppressWarnings({
  seurat_integrated_angl <- integrate_by_features(se,
    angl,
    process = TRUE
  )
})
seurat_integrated_angl
```


## plot integration
```{r}
Seurat::DimPlot(seurat_integrated_angl, reduction = "umap", group.by = "Batch")
Seurat::DimPlot(seurat_integrated_angl, reduction = "umap", group.by = "Group")
```

# Comparison to using HVGs
## using HVGs
### get HVGs
```{r, get HVGs}
se_list <- SplitObject(se, split.by = batch_key)
se_list <- lapply(se_list, NormalizeData)
# anglemania chose only 324 genes based on the thresholds
#   to make it comparable we only choose 324 highly variable genes:
hvg_features <- Seurat::SelectIntegrationFeatures(se_list, nfeatures = 324)
```

### integration
```{r, integration option 2}
suppressWarnings({
  seurat_integrated_hvg <- integrate_seurat_list(se_list,
    features = hvg_features,
    process = TRUE
  )
})
seurat_integrated_hvg
```

### plot integration
```{r}
Seurat::DimPlot(seurat_integrated_hvg, reduction = "umap", group.by = "Batch")
Seurat::DimPlot(seurat_integrated_hvg, reduction = "umap", group.by = "Group")
```

## using 2000 HVGs
### get HVGs
```{r}
hvg_features <- Seurat::SelectIntegrationFeatures(se_list, nfeatures = 2000)
```

```{r}
suppressWarnings({
  seurat_integrated_hvg_2000 <- integrate_seurat_list(se_list,
    features = hvg_features,
    process = TRUE
  )
})
seurat_integrated_hvg_2000
```

### plot integration
```{r}
Seurat::DimPlot(seurat_integrated_hvg_2000, reduction = "umap", group.by = "Batch")
Seurat::DimPlot(seurat_integrated_hvg_2000, reduction = "umap", group.by = "Group")
```

```{r}
sessionInfo()
```