---
title: "Functions_Showcasing"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: "paged"
    toc: true
    toc_depth: 2
    toc_float: true
    theme: simplex
vignette: >
  %\VignetteIndexEntry{Other Functions in anglemania}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# .libPaths(.Library)
```

# Introduction
In this Vignette we want to quickly show the few other functions that are available in 
anglemania besides the two main ones: `create_anglemania_object` and `anglemania`.


```{r, load libraries}
library(anglemania)
library(SingleCellExperiment)
```

# Normal anglemania workflow
```{r}
sce <- sce_example()
batch_key <- "batch"
angl <- create_anglemania_object(sce, batch_key = batch_key)
angl <- anglemania(angl)
angl
```

# under the hood of the anglemania function
* anglemania is run on the anglemania object and it calls three functions:
    * factorise: 
        * creates a permutation of the input matrix to create a null distribution
        * computes the cosine similarity (or spearman coefficient or diem) between gene expresion 
          vector pairs matrix for both the original and permuted matrices
        * computes the zscore of the relationship between the gene pairs taking the mean and standard 
          deviation of the null distribution
        * it does this for every batch in the dataset!
    * get_list_stats
        * computes the mean and standard deviation of the zscores across the matrices from
          the different batches.
    * select_genes
        * filters the gene pairs by the mean zscore and the signal-to-noise ratio (SN, i.e. the 
          mean divided by the standard deviation).

## factorise
```{r}
barcodes_by_batch <- split(rownames(colData(sce)), colData(sce)[[batch_key]])
counts_by_batch <- lapply(barcodes_by_batch, function(x) {
    counts(sce[, x]) %>% sparse_to_fbm()
})
counts_by_batch[[1]][1:10, 1:6]
class(counts_by_batch[[1]])
factorised <- lapply(counts_by_batch, factorise)
factorised[[1]][1:10, 1:6]
```

## get_list_stats
* get_list_stats gets called on the anglemania object so we create and inspect a anglemania object
```{r}
sce <- sce_example()
batch_key <- "batch"
angl <- create_anglemania_object(sce, batch_key = batch_key)
list_stats(angl) # this slot is empty
angl <- anglemania(angl) 
names(list_stats(angl))
class(list_stats(angl))
list_stats(angl)$mean_zscore[1:10, 1:6]
list_stats(angl)$sn_zscore[1:10, 1:6]
```

## select_genes
* under the hood, anglemania calls select_genes on the anglemania object with the default thresholds
  zscore_mean_threshold = 2.5, zscore_sn_threshold = 2.5
* we can use select_genes to change the thresholds without having to run anglemania again
```{r}
previous_genes <- get_anglemania_genes(angl)
angl <- select_genes(angl,
                     zscore_mean_threshold = 2,
                     zscore_sn_threshold = 2)
# Inspect the anglemania genes
new_genes <- get_anglemania_genes(angl)

length(previous_genes)
length(new_genes)
```